Bias variance trade-off comparison of regression models

### 2022-11-24 doc/paper.tex
Final improvements

\section{Abstract}
Comparing the different regression methods of OLS and ridge regression, we found the regularization provided by ridge regression to decrease our MSE by 14\% by studying the bias variance trade-off of these methods. Regularization was sufficient for delaying overfitting, but a too large regularization value could increase the MSE unnecessarily. Furthermore, decision trees yielded poor results for our regression problem, more than doubling the MSE of our results found by ridge regression. We turned to neural networks, which were also unable to beat the results obtained with ridge regression even when using a very complex architecture consisting of 5 hidden layers. Ridge regression took the crown with an MSE of 0.0031.

@@

This sections covers the theoretic background which underlies the methods used in this paper.

@@

A common problem with OLS is what is known as overfitting (we will discuss this in greater detail in later sections). A symptom of overfitting is that the values in the parameter vector $\bm{\beta}$ grow very large, and vary wildly for small changes in our dataset. To counteract this, we can add to the cost function a term which increases with the values in $\bm{\beta}$. Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also minimizing the absolute value of the parameters. We now define a new cost function

@@

Mathematically, each node on a layer is a linear combination of all the nodes of the previous layer, fed into the activation function for the current layer.

\subsubsection{Decision trees}
A decision tree will split our dataset, the predictor space, into different non-overlapping areas for which we will make the yield the same prediction for every observation that falls into this area. The prediction will be the average of the observations present in the area during training. The areas, or regions, are chosen to minimize the MSE. The decision tree has an advantage of being very intuitive and explainable, as the observations are placed into these regions using simple conditions.

@@

Secondly, we will perform our linear regression on this new dataset and store the predictions of the test data. Thirdly, we repeat for a given large number of iterations. Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem \cite{CLT} obtain a normal distribution of these performance metrics. The law of large numbers \cite{LLN} then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance. This is dependent on our assumption that our data is independently and identically distributed.

@@

To obtain a valid comparison accross the various regression methods, we used methods found in scikit learn's machine learning library such as  LinearRegression, Ridge, MLPRegressor and DecisionTreeRegressor. We specify the model we wish to use such that it outputs a mean squared error as it increases in complexity. The complexity for OLS and ridge is the polynomial degree up to which we create the design matrix, the model complexity for the feed forward neural network is the amount of hidden nodes in our hidden layers, and the complexity in the decision tree is represented by its depth. For the feed forward neural network we implemented both one and two hidden layers, and distributed the hidden nodes equally across the layers. The feed forward neural network is trained for 2000 epochs using the logistic activation function and the adam scheduler.

@@

\subsection{Datasets}

Below is a short description of each dataset used in this report, along with a table (Table.~\ref{datasettable}) listing their characteristics.

@@

The Franke Function is a simple weighted sum of exponentials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems. For OLS and ridge regression, our training data will be based on a 9th degree polynomial fit of the Franke Function, with 56 features. For the feed forward neural network and decision tree, we will use the x and y values directly. The Franke Function is defined as follows.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=5in]{figs/BV_OLS.png}}
\caption{Bias variance trade-off for ordinary least squares linear regression. Y axis represents model complexity as polynomial degrees.}
\label{OLS}
\end{figure}

@@

For the feed forward neural network, we can see from (Fig.~\ref{ffnn1}) that a FFNN with one hidden layer will overfit when the hidden layer contains over 80 hidden nodes. The test MSE reaches a minimum of 0.022, a whole order of magnitude worse than our previous results. Unsatisfied with this result, we turn to an architecture with more hidden layers in a persuit of a lower MSE. Strangely, the cause of the MSE increasing for hidden node counts greater than 80 is a large increase in the bias. Typically, the bias will start high and go downwards as the model complexity increases. This might be because the FFNN simplifies the relationship learned to describe the pattern, increasing the bias. The variance does also increase however, but not as much as the bias.

@@

\subsection{Conclusion}
Ridge proved to be the best regression model, as the regularization allows us to drive down the MSE to an impressive 0.0031 whilst delaying the overfitting. Ridge was able to mentain a low bias and variance, an impressive 14\% decrease from OLS' results. Decision trees didn't overfit, but was unable to beat the MSE of ridge and OLS as the bias variance trade-off featured both a higher bias and variance. The FFNN required a great increase in computational cost whilst not being able to beat the MSE of OLS and Ridge. For 5 hidden layers, the FFNN yields a lower MSE than the decision tree.

@@

\begin{table*}[h]
\caption{This table shows the file which created every plot in the report}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l}
Figure & Shell command \\
\hline
\ref{OLS} & \texttt{python3 OLS\_bias\_variance.py}\\
\ref{Ridge} & \texttt{python3 ridge\_bias\_variance.py}\\
\ref{tree} & \texttt{python3 decision\_tree\_bias\_variance.py}\\
\ref{ffnn1} & \texttt{python3 ffnn\_bias\_variance.py}\\
\ref{ffnn2} & \texttt{python3 ffnn\_bias\_variance.py}\\
\hline
\end{tabular}
\end{center}
\end{table*}

>>>>

\section{Abstract}

We compared linear regression, Neural Networks and Random Forests in their ability to fit the Franke Function without overfitting. This was done by comparing the bias variance tradeoff for each of the models. Comparing the different regression methods of OLS and ridge regression, we found the regularization provided by ridge regression decreased our MSE by $14\%$, providing both a low bias and variance. Regularization was sufficient for delaying overfitting, but a too large regularization value could increase the bias unnecessarily. Furthermore, decision trees yielded poor results for our regression problem, more than doubling the MSE of our results found by ridge regression and exhibiting high variance. We turned to neural networks, which were also unable to beat the results obtained with ridge regression even when using a very complex architecture consisting of 5 hidden layers. Ridge regression took the crown with an MSE of $0.0031$.

@@

This section covers the theoretic background which underlies the methods used in this paper.

@@

A common problem with OLS is what is known as overfitting. A symptom of overfitting is that the values in the parameter vector $\bm{\beta}$ grow very large, and vary wildly for small changes in our dataset. To counteract this, we can add to the cost function a term which increases with the values in $\bm{\beta}$. Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also minimizing the absolute value of the parameters. We now define a new cost function

@@

\subsubsection{Decision trees}
A decision tree will split our dataset, the predictor space, into different non-overlapping areas for which we will yield the same prediction for every observation that falls into this area. The prediction will be the average of the observations present in the area during training. The areas, or regions, are chosen to minimize the MSE. The decision tree has an advantage of being very intuitive and explainable, as the observations are placed into these regions using simple conditions.

@@

Secondly, we will perform our regression on this new dataset and store the predictions of the test data. Thirdly, we repeat for a given large number of iterations. Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem \cite{CLT} obtain a normal distribution of these performance metrics. The law of large numbers \cite{LLN} then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance. This is dependent on our assumption that our data is independently and identically distributed.

@@

To obtain a valid comparison accross the various regression methods, we used methods found in scikit learn's machine learning library such as  LinearRegression, Ridge, MLPRegressor and DecisionTreeRegressor. We specify the model we wish to use such that it outputs a mean squared error as it increases in complexity. The complexity for OLS and ridge is the polynomial degree up to which we create the design matrix, the model complexity for the feed forward neural network is the amount of hidden nodes in our hidden layers, and the complexity in the decision tree is represented by its depth. For the feed forward neural network we implemented both one and five hidden layers, and distributed the hidden nodes equally across the layers. The feed forward neural network is trained for 2000 epochs using the logistic activation function and the adam scheduler.

@@

@@

The Franke Function is a simple weighted sum of exponentials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems.

@@

\begin{figure}[h]
\centerline{\includegraphics[width=5in]{figs/BV_OLS.png}}
\caption{Bias variance trade-off for ordinary least squares linear regression. Y axis represents model complexity as polynomial degrees.}
\label{OLS}
\end{figure}

@@

For the feed forward neural network, we can see from (Fig.~\ref{ffnn1}) that a FFNN with one hidden layer will overfit when the hidden layer contains over 80 hidden nodes. The test MSE reaches a minimum of 0.022, a whole order of magnitude worse than our previous results. Strangely, the cause of the MSE increasing for hidden node counts greater than 80 is a large increase in the bias. Typically, the bias will start high and go downwards as the model complexity increases. This might be because the FFNN simplifies the relationship learned to describe the pattern, increasing the bias. The variance does also increase however, but not as much as the bias. Unsatisfied with this result, we turn to an architecture with more hidden layers in a pursuit of a lower MSE.

@@

\subsection{Conclusion}
Ridge proved to be the best regression model, as the regularization allows us to drive down the MSE to an impressive 0.0031 whilst delaying the overfitting. Ridge was able to mentain a low bias and variance, an impressive 14\% decrease from OLS' results. Decision trees didn't overfit, but were unable to beat the MSE of ridge and OLS as the bias variance trade-off featured both a higher bias and variance. The FFNN required a great increase in computational cost whilst not being able to beat the MSE of OLS and Ridge. For 5 hidden layers, the FFNN yields a lower MSE than the decision tree.

@@

\begin{table*}[h]
\caption{This table shows the file which created every plot in the report}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l}
Figure & Shell command \\
\hline
\ref{OLS} & \texttt{python3 OLS\_bias\_variance.py}\\
\ref{Ridge} & \texttt{python3 ridge\_bias\_variance.py}\\
\ref{tree} & \texttt{python3 decision\_tree\_bias\_variance.py}\\
\ref{ffnn1} & \texttt{python3 ffnn\_bias\_variance.py}\\
\ref{ffnn5} & \texttt{python3 ffnn\_bias\_variance.py}\\
\hline
\end{tabular}
\end{center}
\end{table*}
